Efficiency

### 2018-01-21 DraftExec.Rnw
axial mixed update, and john and jim edits

It is worthwhile explaining briefly the origins and intent of this paper, how it came to be written, and how it ought to be read.

The purpose is to investigate the suitability of the pump-jet selection for Australia's future fleet of submarines.  Amongst this considerable controversy surrounding the submarines, this particular technology choice deserves special attention.  Much has been made of the significance of the pump-jet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pump-jet rendered propellers obsolete \parencite{naval2018}, and features extremely prominently in the all the marketing and literature. It seems to be accepted knowledge that the proposal from Naval Group (formerly DCNS) was selected because of its acoustic superiority \parencite{stewart2016}, with the pumpjet apparently comprising the jewel in the crown.

Having done my higher education in physics, and spent considerable time considering military systems running a small start-up defence company (Rubber Ducky Defence) I found the pumpjet choice particulaly curious.  As is commonly quipped, including in defence circles, it can reliably assumed when it comes to technology choices "there's no free lunch in physics".  Whilst all technologies steadily improve, and new ones continuously emerge, it's inevitably the case that significant technology switches are made, some trade-offs are necessary. In the case of the pumpjets, the trade-offs were obvious: propulsive efficiency.

It's definitely true that efficiency isn't everything in military technology.  For Rubber Ducky, I spent considerable time and money designing a small speed-boat and a medium-weight torpedo for naval combat, and in both cases I selected a waterjet or pumpjet for those purpose. Pumpjets, or some kind of "decelerating ducted propeller" (explained in detail later) suited very well the sorts of speeds and operational circumstances that they were required to operate in, which were well over 30kt. However, having made that selection, and on several occasions wrestled with adapting the boat to alternative purposes, (such as mine warfare, including the towing of heavy sweeps and decoys) I found myself continually balancing the inclination to design a better, more specialised tool for each job, with the desire to compromise performance consolidate capability into a multi-role, more flexible capability.  The trade-offs inherent in propulsion systems, and the relative efficency of jets at high speeds and propellers at low speeds was a familiar encounter.

The selection of the DCNS bid to design the future submarine attracted considerable attention and controversy on a number of fronts. The choice of the pumpjet, and the question of efficiency at low speeds, has been touched on briefly in various informed discussions \parencite{stanford2017} \parencite{davies2017}.  However, it has remained the case that even extremely well-informed commentators reluctant to quantify and qualify the plausible scale of this trade-off, and what it might mean strategically and tactically for the submarine. Everyone knew it was an issue, but no-one confident enough in the scale (or even direction \parencite{davies2017}) of the impact to make it clear whether it mattered. Even the head of the progam to build the submarine acknowledged an improvement in dived endurance might be possible with conventional propellers \parencite{sbs2017}, which seemed a clear acknowledgment that some trade-off was acknowledged, even by the strongest advocates of the pumpjet.

The development of the full substance of this paper has come about due to the determination and interest of Gary Johnson.  Through mutual networks in industry and academia we discovered that we had both been intrigued by the technology choice, and considered it of potential significance to Australia's national interest. Australia is fortunate to have citizens so concerned with national interest to so substantially invest in research to advance public debate.  Unless he had commissioned me to write it, there is no way that I would have been able to make the sustained and deliberate research which the topic required to reach a useful level of quantification and confidence to meaningfully inform debate.

I should hasten to add that my expertise in this area is derived largely from an academic background which a different field of physics (particle physics), which enhanced only my aptitude for understanding and analysing physical concepts, paired with a substantial exposure to the relevant disciplines and industry from a commercial perspective as an entrepreneur developing new products.  I don't claim to be an authority, but reasonably capable of grasping the works and words of those who are.

@@

It should be noted that the technical fields involved here are vast and rapidly evolving.  It is quite possible that there are other effects and phenomena which could be relevant which aren't mentioned here, and it's also true that additional advances in the state of the art are being made continually, and might not be known about in public literature.

However, the nature of scientific discovery and technical progress is that many of best discoveries are actually built-upon and confirmed, rather than swept aside, by later developments.  As such, plenty of foundational principals, in particular the conservation laws of energy and momentum and essential principals of mechanics and thermodynamics are just as true, right, and relevant as they were hundreds of years ago.  Consequently, whilst we cannot always know the precise degree of advance in a particular field of engineering, we can still confidently know some of the fundamental bounds and constraints within which that field will operate.

@@

The selection of the pumpjet propulsion system is quite probably the most significant new technology selection for Australia's future submarines.  Unlike other new technolgoies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other conventionally powered submarines have adopted pump-jet technology, which is currently only used by nuclear submarines.

@@

The comparison between two otherwise identical submarines (though not necessarily limited to engaging just each other in combat), provides a useful and suitable framework for evaluating a single significant technology choice, where most other significantly novel or new technologies were apparently excluded from consideration \parencite{stanford2017}.  The conclusion of this report is that the following statement can be made with confidence:
\\
\\
\emph{In a comparison between two otherwise identical submarines, the one with the pump jet will alwayas have a lower dived endurance, a lower dived range, a worse indescretion ratio, a lower overall endurnace, and a lower overall range, than the one with a conventional propeller.  This will confer a substantial tactical and strategic advantage on the conventionally propelled submarine in a very broad range of operational scenarios.}
\\
\\
Whilst it is broadly acknowledged that propellers have superior efficiency at lower speeds, but jets might have higher efficiency at higher speeds, the precise point of cross-over seemed unclear in public debate.  One informed commentator suggested that the crossover point was likely to be below or near the transit speed of the submarine, and that the pumpjet may offer enhanced range and endurance over the whole mission profile as a consequence \parencite{davies2017}.  This report finds no evidence of a pumpjet of the type used on submarine every being recommended as superior for efficiency in general, and even closely related technologies for more commecial use not advocated below 25kt. During my research I encountered domain experts confident that a pumpjet could simply not feasibly be superior to a suitable propeller at speeds below 18kt.  Evidence of propellers remaining generally superior to waterjets and pumpjets in the speed ranges up to and even over 30kt is abundant.  Consequently, the possibility of some cross-over point leading to any kind of range or endurance advantage relevant for a conventional submarine can be safely ruled out.

This report also explores in some detail the theoretical concepts which drive the limitations of pumpjet performance at lower speeds in comparison to propellers.  Enduring and fundamental principals of momentum theory provide robust limitations on the circumstances where a propeller's 'ideal efficiency' might be improved by the addition of a duct/or shroud, one type of which constitutes the transition to a 'pumpjet'.  These principals stronly support the the notion that pumpjets will not and indeed cannot suddenly overcome their efficiency disadvantage at low speeds by means of technical ingenuity or innovation.

Furthmore, a particular hydrodynamic phenonomon called 'flow separation' can be identified as the primary cause of the pumpjet's declining efficiency at very low speeds. Put simply, the duct provides an additional hurdle or barrier though which the water must pass when moving through the propuslor.  At very low speeds waterspeeds, this additional resistance is sufficient to stop some of the flow progressing through the jet, creating unsteady eddies and vortices which sap energy from the system, but produce no thrust, driving down efficiency to zero at very low flow-rates.

The design of the pumpjet for military purposes, namely the reduction of acoustic signatures effectively mandates that this must be the way that pumpjets work.  The act of slowing the water down is necessarily equivalent to the act of raising its pressure.  It is this elevated pressure at the point where the blades to their work on the water which enables the blades to avoid cavitation (the rapid expansion and collapse of small voids or bubbles in the water) which is very large source of noise underwater.  Maintaing this advantage to the same level necessitates a duct design which slows water down to the same extent.  The act of slowing water down when it's already going slowly is what causes some of it to stop, which is when flow separation occurs.

@@

The model has been made available publically in the form of a web-app, so that alternative assumptions, including alternative efficiency curves may be tested by the general public.

The model in its central assumptions was able to achieve very realistic results, which were consistent with other public estimates of performance parameters of conventional submarines and known characteristics \ref{buckingham2008submarine}.  A variety of different assumptions regarding the hotel load, battery density and different combinations of efficiency curves were tested.  The central assumption values showed that a submarine with a conventional propeller had a very substantial advantage in terms of range and endurance across the entire speed range.  This advantage reached a proportional maxium at around 6kt of over 60\%.  However, in nominal terms, the greatest differences occurred at slightly lower speeds.  Under central assumptions, the propeller submarine had an advantage in submerged range of some 200nm at 5kt, and approximately a two-day advantage in dived endurance at 4kt.

@@

The model was used to test a wide range of different assumptions.  A further key result of the study established the sensitity of the difference to different assumptions regarding hotel load, and battery energy density.  This showed that the difference between the two systems was made substantially larger with smaller hotel loads, and larger stored electical energy (as might be achieved with lithium-ion batteries), and these two effects combined to give some particularly startling results.  Circumstances where the \textit{difference} in dived range and endurance between the two systems in terms of range and endurance exceeded the \textit{entire} dived range and endurance achieved by many modern conventional submarines frequently arose.

The model demonstrated clearly that all of the most pronounced advantages in dived range and endrance occurred at speeds around 3-7kt, generally within plausible speeds achievable with modern AIP systems.  The addition of AIP was briefly considered which revealed that a difference in dived endurance of the order of one month, or thousands of nautical miles, would likely have emerged between the two propulsion systems.

The consequences of flow separation also has acoustic consequences, which this report explores.  The vortexes and eddies which are created by flow separation tend to be unsteady, and shed periodically into the main flow.  This results in pressure fluctations inside the shroud and more intense, unsteady large-scale types of turbulence occurring as these vortices are ingested into the rotor blades.  These effects are are likely to cause some accoustic signature which becomes lower in frequence as the submarine slows down further and the vortexes become larger, and shed more slowly.  This is in contrast to early stages of cavitation, which would be high frequency at when bubble sizes are small, and move to lower fequences as speed increases and bubbles expand.  Lower frequencies are of particular significance because they tend to propagate further, and would be less effectively shielded by the duct.

It remains plausible that in the absence of cavitation the noise generated in the propulsor by such flow separation is not loud enough to be of significant tactical concern. However, it is extremely unlikely that at some very low speeds, where conventional propellers experience no cavitation and enjoy steady, smooth flows over the blades, that a pumpjet could actually have a lower acoustic signature, even in terms of radiated noise.  Flow separtion demand that two or three times as much energy is wasted by a pumpjet creating and destroying these unsteady vortexes than a propeller would require to create equal thrust. Consequently, the claim that pumpjets are generally acoustically superior should be treated with some caution.  This claim has strong grounds wherever a conventional propeller might experience cavitation, such as at higher speeds.  But at some very low speeds it is unlikely to be true.

@@

This effect is of profound importance to understanding the operations of conventional submarines, which have extremely constrained energy stores when operating under the surface. Diesel fuel when burned has an energy density of approximately 45MJ/kg.  In contrast, a lead-acid battery might have energy densities in the range of 0.08-0.14MJ/kg. With something like 400 times as much energy per kg embarked in diesel, operating on batteries imposes an extreme demand for economy on propulsive power.  Happily for submarines, the square law for drag, and cubed law for power, allow an almost commensurate reduction in power demands to take place by slowing down to very slow speeds when submerged. Hypothetically, a conventional submarine might have a total range of 10,000nm from using it's diesel payload at 8kt.  If an equal weight of lead-acid batteries as fuel were carried, the submarine could only travel about 25nm when submerged at the same speed in a single charge.  But by travelling at 4kt, that quickly increases to 100nm, and 400nm at 2kt (neglecting hotel load here for simplicity).

@@

One more plausible argument is worth consideration, namely that a safety requirement to be able to operate on batteries without the reactor drives a need to minimise hotel load. For all the reasons discussed above, it's completely implausible that the reactor would ever be shut down volunarily during operations.  One would think that the sorts of emergencies or accidents which would cause a reactor are likely to also mean that the reactor can't be restarted.  However, it is plausible that through the development of some instability, perhaps through an unknown fault or dramatic power reductions of the type which I described should be avoided, it might be the case that the reactor would need to be shut down, but could plausibly be restarted.  However, as discussed above, the restart would need to occur either within minutes, or probably in a couple of days.

@@

\subsection{Ducted Propellers}
A ducted propeller is a propeller in any kind of tubular shroud, casing or nozzle.  Ducted propellers fall into two categories, one of which is referred to frequently and interchangeably as a pump-jet.  The two types are called the accelerating type (also known as a Kort Nozzle, after a company that made them popular in the 20th Century) and the decelerating duct.  The names describe the different effects that each type of nozzle have on the flow speed of the water at the point where it meets the propeller (which is also called an impeller, particularly when associated with decelerating ducts, or pumpjets).

@@

The addition of some new parts such as stator blades (to straighten out the flow) and other possible features which make the propeller system more like pump or turbomachinery components might be associated with systems more commonly referred to as pumpjets than ducted propellers.  Not all decelerating ducts have stators, but the impact of the duct on the stream will remain. The key concepts behind the operation of these systems, however, remains essentially similar in their nature and design intent.  In the presence of a duct propellers might be advantageously shaped differently, including maintaing a larger chord lenght right to the shroud, rather than tapering as open propellers tend to lessen vorexs and cavitation on the extremities.  However, there is not to my knowledge any single defining physical characteristic which the transition to the name 'impeller' implies, which woud not be effectively equivalent to specialised propellers for decelerating ducts.  In systems tending to be described as pumpjets or waterjets hubs tend to be larger, blade-numbers higher, twist, skew and rake lower.  No definitive transition point seems to be accepted as having great significance.  Most commonly literature will refer to a 'rotor', a more general term which could encompass both propellers and impellers.

@@

@@

\subsection{Recent Commentary}
There has been already some attention given to the probable efficiencies of pump jets at low speed. Recently Andrew Davies published a piece in the The Strategist (ASPI) arguing that the pump jet may actually save fuel at transit speeds (around 8kt) and overall improve the endurance of the submarine, despite likely having lower efficiency at low speeds such as patrol speed \parencite{davies2017}.  This argument relies on pump-jet efficiency still overtaking that of propellers at what is considered a relatively low speed for many applications.  As Andrew acknowledges, further quantification of the likely efficiency is required, as the particular chart which he cites is not quantified.  This section aims to identify as far as is reasonably possible from public literature how the efficiency of a likely pump-jet for a submarine might perform in quantitative terms, relative to a propeller.

@@

\begin{figure}[h]
\captionsetup{width=1\textwidth}
\includegraphics[width=1\textwidth]{JetHighFlow.png}
\caption{At high flow rates the water has sufficient velocity (indicated by arrow length) to overcome the adverse pressure gradients created by the duct shape, shown by the change from green to red. After the impeller has added further pressure to the water, the water accelerates out of the contracting nozzle through the favourable pressure gradient (red to green) to produce a jet, reaching equal pressure to outside at the exit}
\label{fig:JetHighFlow.png}
\end{figure}

@@

\begin{figure}[h]
\captionsetup{width=1\textwidth}
\includegraphics[width=1\textwidth]{JetLowFlow.png}
\caption{At low flow rates the water has insufficient velocity to overcome the adverse pressure gradients created by the duct shape, and flow separation begins to occur on the duct intake, and between the blades beginning near the roots (stall).  Flow through the jet would be unsteady, with excessive energy expended creating vortices and turbulence that results in little thrust.}
\label{fig:JetLowFlow.png}
\end{figure}

@@

Given the fundamental nature of the flow-rate and intake area to the design of pump-jets, and the great lenght of time for which they have been understood and discussed as a crucial design input for a waterjet, it is altogether implausible that some large increase in intake area can be achieved without incurring necessary and unavoidable tradeoffs.

>>>>

The purpose of this paper is to investigate the suitability of the pumpjet propulsor technology selection for Australia's future fleet of submarines.  Amongst this considerable controversy surrounding the submarines, this particular technology choice deserves special attention.  Much has been made of the significance of the pumpjet in the DCNS (now Naval Group) bid for the submarine, to the extent that it has been claimed that the pumpjet rendered propellers obsolete \parencite{naval2018}, and features extremely prominently in the all the marketing and literature. It seems to be accepted knowledge that the proposal from Naval Group (formerly DCNS) was selected because of its acoustic superiority \parencite{stewart2016}, with the pumpjet apparently comprising the jewel in the crown.

The selection of the DCNS bid to design the future submarine attracted considerable attention and controversy on a number of fronts. The choice of the pumpjet, and the question of efficiency at low speeds, has been touched on briefly in various informed discussions \parencite{stanford2017} \parencite{davies2017}.  However, it has remained the case that even extremely well-informed commentators reluctant to quantify and qualify the plausible scale of this trade-off, and what it might mean strategically and tactically for the submarine. Everyone knew it was an issue, but no-one confident enough in the scale of the impact (or even at transit speeds the direction of the impact \parencite{davies2017}) to make it clear whether it mattered. Even the head of the progam to build the submarine acknowledged an improvement in dived endurance would be possible with conventional propellers \parencite{sbs2017}, which seemed a clear acknowledgment that some trade-off was acknowledged, even by the strongest advocates of the pumpjet.  But despite this, the chiefs of DCNS and SEA 1000 both remain adamant that the pumpjet was an "integral part of our bid" \parencite{guillou2017} and that it can be efficient "across the entire speed range" \parencite{sammut2017}.

Having done my higher education in physics, and spent considerable time considering military maritime technology running a small start-up defence company I found seemingly unresolved series of exchanges regarding the pumpjet choice particulaly curious.  As is commonly quipped, including in defence circles, it can be reliably assumed when it comes to technology choices "there's no free lunch in physics".  Whilst all technologies steadily improve, and new ones continuously emerge, it's inevitably the case that when significant technology switches are made, some trade-offs are necessary. In the case of the pumpjets, the trade-offs what might have been expected appeared obvious, and probably unavoidable to some extent to all those experienced in maritime systems: it would have lower propulsive efficiency, particularly at low speeds.

It's definitely true that efficiency isn't everything in military technology.  In recent years I was heavily involved in designing a small speed-boat and a medium-weight torpedo for naval combat, and in both cases I selected a waterjet or pumpjet for those purpose. Pumpjets, or some kind of "decelerating ducted propeller" (explained in detail later) suited very well the sorts of speeds and operational circumstances that they were required to operate in, which were well over 30kt. However, having made that selection, and on several occasions wrestled with adapting the boat to alternative purposes, (such as mine warfare, including the towing of heavy sweeps and decoys) I found myself continually balancing the inclination to design a better, more specialised tool for each job, with the desire to compromise performance consolidate capability into a multi-role, more flexible capability.  The trade-offs inherent in propulsion systems, and the relative efficency of jets at high speeds and propellers at low speeds was a familiar encounter.

The development of this line of inquiry into of the full substance of this paper has come about not due to my own intrigue, but largely due to the determination and interest of Gary Johnson.  Through mutual networks in industry and academia we discovered that we had both been intrigued by the technology choice, and considered it of potential significance to Australia's national interest. Australia is fortunate to have citizens so concerned with national interest to so substantially invest in research to advance public debate.  Unless he had commissioned me to write it, there is no way that I would have been able to make the sustained and deliberate inquiry which the topic required to reach a useful level of quantification and confidence to meaningfully inform debate.

I should hasten to add that my expertise in this area is derived largely from an academic background whithin a different field of physics (particle physics), which enhanced only my aptitude for understanding and analysing physical concepts, paired with a substantial exposure to the relevant disciplines and industry from a commercial perspective as an entrepreneur developing new products.  I don't claim to be an authority, but reasonably capable of grasping the works and words of those who are.

@@

It should be noted that the technical fields involved here are vast and rapidly evolving.  It is quite possible that there are other effects and phenomena which could be relevant which aren't mentioned here, and it's also true that additional advances in the state of the art are being made continually, and might not be known about in public literature.  If other subject matter experts have additional or better insights into the topic than those presented here, I would welcome their further comment and contribution to debate.

However, the existence of unknown developments and advances does not mean that realistic approximate bounds and limits of technologies cannot be found from public inquiry.  Scientific discovery and technical tends to confirm many of the best discoveries are actually built-upon and confirmed, rather than simply sweep them aside  As such, plenty of foundational principals, in particular the conservation laws of energy and momentum and essential principals of mechanics and thermodynamics are just as true, right, and relevant as they were hundreds of years ago.  Consequently, whilst we cannot always know the precise degree of advance in a particular field of engineering, we can still confidently know some of the fundamental bounds and constraints within which that field will operate.  My research into this topic has confirmed that this is certainly the case in the relevant arts of hydrodynamics and turbomachinery, where a number of canonical works from the middle of last century are still very frequently referenced in the most modern literature for outining the foundational principals.

@@

The choice of the pumpjet propulsion system is quite probably the most significant new technology inclusion for Australia's future submarines.  Unlike other modern technolgoies such as Air Independent Propulsion and Lithium Ion Batteries, which are either in service or entering service in multiple other navies, no other navies have have adopted pumpjet technology for their conventional sumbarines, which is currently only used by nuclear submarines.

@@

The comparison between two otherwise identical submarines (though not necessarily limited to engaging just each other in combat), provides a useful and suitable framework for evaluating a single significant technology choice, where most other significantly novel or new technologies were apparently excluded from consideration \parencite{stanford2017}.  The conclusion of this report is that the following statement can be made with confidence:
\\
\\
\emph{In a comparison between two otherwise identical submarines, the one with the pump jet will always have a lower dived endurance, a lower dived range, a worse indiscretion ratio, a lower overall endurace, and a lower overall range, than the one with a conventional propeller.  This will confer a substantial tactical and strategic advantage on the conventionally propelled submarine in a very broad range of operational scenarios.}
\\
\\
Whilst it is broadly acknowledged that propellers have superior efficiency at lower speeds, but jets might have higher efficiency at higher speeds, the precise point of cross-over seemed unclear in public debate.  One informed commentator suggested that the crossover point was likely to be below or near the transit speed of the submarine, and that the pumpjet may offer enhanced range and endurance over the whole mission profile as a consequence \parencite{davies2017}.  This report finds no evidence of a pumpjet of the type used on submarine ever being recommended as superior for efficiency in general, and even closely related technologies for more commecial use not advocated below 25kt. During my research I encountered domain experts confident that a pumpjet could simply not feasibly be superior to a suitable propeller at speeds below 18kt.  Evidence of propellers remaining generally superior to waterjets and pumpjets in the speed ranges up to and even over 30kt is abundant.  Consequently, the possibility of some cross-over point leading to any kind of range or endurance advantage relevant for a conventional submarine can be safely ruled out.  The basis for claims made by public officials to the contrary, such as those by Rear Admiral Sammut \parencite{sammut2017}, must be called into question.

This report also explores in some detail the theoretical concepts that drive the limitations of pumpjet performance at lower speeds in comparison to propellers.  Enduring and fundamental principals of momentum theory provide robust limitations on the circumstances where a propeller's 'ideal efficiency' might be improved by the addition of a duct/or shroud, one type of which constitutes the transition to a 'pumpjet'.  These principals stronly support the the notion that pumpjets will not and indeed cannot suddenly overcome their efficiency disadvantage at low speeds by means of technical ingenuity or innovation.

Furthmore, a particular hydrodynamic phenonemon called 'flow separation' can be identified as the primary cause of the pumpjet's declining efficiency at very low speeds. Put simply, the duct provides an additional hurdle or barrier though which the water must pass when moving through the propulsor.  At very low waterspeeds, this additional resistance is sufficient to stop some of the flow progressing through the jet, creating unsteady eddies and vortices which sap energy from the system, but produce no thrust, driving down efficiency to zero at very low flow-rates.

The design of the pumpjet for military purposes, namely the reduction of acoustic signatures effectively mandates that this must be the way that pumpjets work.  The act of slowing the water down is necessarily equivalent to the act of raising its pressure.  It is this elevated pressure at the point where the blades to their work on the water which enables the blades to avoid cavitation (the rapid expansion and collapse of small voids or bubbles in the water) which is very large source of noise underwater.  Maintaining this advantage to the same level necessitates a duct design which slows water down to the same extent.  The act of slowing water down when it's already going slowly is what causes some of it to stop, which is when flow separation occurs.

@@

The model has been made available publically in the form of a web-app, so that alternative assumptions, including alternative efficiency curves may be tested by the interested parties.

The model in its central assumptions was able to achieve very realistic results, which were consistent with other public estimates of performance parameters of conventional submarines and their known characteristics \ref{buckingham2008submarine}.  A variety of different assumptions regarding the hotel load, battery density and different combinations of efficiency curves were tested.  The central assumption values showed that a submarine with a conventional propeller had a very substantial advantage in terms of range and endurance across the entire speed range.  This advantage reached a proportional maximum at around 6kt of over 60\%.  However, in nominal terms, the greatest differences occurred at slightly lower speeds.  Under central assumptions, the propeller submarine had an advantage in submerged range of some 200nm at 5kt, and approximately a two-day advantage in dived endurance at 4kt.

@@

The model was used to test a wide range of different assumptions.  A further key result of the study established the sensitivity of the difference to a range of assumptions regarding hotel load, and battery energy density.  This showed that the difference between the two systems was made substantially larger with smaller hotel loads, and larger stored electical energy (as might be achieved with lithium-ion batteries), and these two effects combined to give some particularly startling results.  Circumstances where the \textit{difference} in dived range and endurance between the two systems in terms of range and endurance exceeded the \textit{entire} dived range and endurance achieved by many modern conventional submarines frequently arose.

The model demonstrated clearly that all of the most pronounced advantages in dived range and endurance occurred at speeds around 3-7kt, generally within plausible speeds achievable with modern AIP systems.  The addition of AIP was briefly considered which revealed that a difference in dived endurance of the order of one month, or thousands of nautical miles, would likely have emerged between the two propulsion systems.

The consequences of flow separation also has acoustic consequences, which this report explores.  The vortexes and eddies which are created by flow separation tend to be unsteady, and shed periodically into the main flow.  This results in pressure fluctuations inside the shroud and more intense, unsteady large-scale types of turbulence occurring as these vortices are ingested into the rotor blades.  These effects are are likely to cause some accoustic signature which becomes lower in frequency as the submarine slows down further and the vortexes become larger, and shed more slowly.  This is in contrast to early stages of cavitation, which would be high frequency at when bubble sizes are small, and move to lower fequencies as speed increases and bubbles expand.  Lower frequencies are of particular significance because they tend to propagate further, and would be less effectively shielded by the duct.

It remains plausible that in the absence of cavitation the noise generated in the propulsor by such flow separation is not loud enough to be of significant tactical concern. However, it is extremely unlikely that at some very low speeds, where conventional propellers experience no cavitation and enjoy steady, smooth flows over the blades, that a pumpjet could actually have a lower acoustic signature, even in terms of radiated noise.  Flow separation demands that two or three times as much energy is wasted by a pumpjet creating and destroying these unsteady vortexes than a propeller would require to create equal thrust. Consequently, the claim that pumpjets are generally acoustically superior should be treated with some caution.  This claim has strong grounds wherever a conventional propeller might experience cavitation, such as at higher speeds.  But at some very low speeds it is unlikely to be true.

@@

This effect is of profound importance to understanding the operations of conventional submarines, which have extremely constrained energy stores when operating under the surface. Diesel fuel when burned has an energy density of approximately 45MJ/kg.  In contrast, a lead-acid battery might have energy densities in the range of 0.08-0.14MJ/kg. With something like 400 times as much energy per kg embarked in diesel, operating on batteries imposes an extreme demand for economy on propulsive power.  Fortuitously for submarines, the square law for drag, and cubed law for power, allow an almost commensurate reduction in power demands to take place by slowing down to very slow speeds when submerged. Hypothetically, a conventional submarine might have a total range of 10,000nm from using it's diesel payload at 8kt.  If an equal weight of lead-acid batteries as fuel were carried, the submarine could only travel about 25nm when submerged at the same speed in a single charge.  But by travelling at 4kt, that quickly increases to 100nm, and 400nm at 2kt (neglecting hotel load here for simplicity).

@@

One more plausible argument is worth consideration, namely that a safety requirement to be able to operate on batteries without the reactor drives a need to minimise hotel load. For all the reasons discussed above, it's completely implausible that the reactor would ever be shut down volunarily during operations.  One would think that the sorts of emergencies or accidents which would cause a reactor are likely to also mean that the reactor can't be restarted.  However, it is plausible that through the development of some instability, perhaps through an unknown fault or dramatic power reductions of the type which I described should be avoided, it might be the case that the reactor would need to be shut down, but could plausibly be restarted.  However, as discussed above, the restart would need to occur either within minutes, or probably wait for a couple of days.

@@

\subsection{Ducted Propellers}
A ducted propeller is a propeller in any kind of tubular shroud, casing or nozzle.  Ducted propellers fall into two categories, one of which is referred to frequently and interchangeably as a pumpjet.  The two types are called the accelerating type (also known as a Kort Nozzle, after a company that made them popular in the 20th Century) and the decelerating duct.  The names describe the different effects that each type of nozzle have on the flow speed of the water at the point where it meets the propeller (which is also called an impeller, particularly when associated with decelerating ducts, or pumpjets).

@@

The addition of some new parts such as stator blades (to straighten out the flow) and other possible features which make the propeller system more like pump or turbomachinery components might be associated with systems more commonly referred to as pumpjets than ducted propellers.  Not all decelerating ducts have stators, but the impact of the duct on the stream will remain. The key concepts behind the operation of these systems, however, remains essentially similar in their nature and design intent.  In the presence of a duct propellers might be advantageously shaped differently, including maintaing a larger chord lenght right to the shroud, rather than tapering as open propellers tend to in order to lessen vortices and cavitation on the extremities.  However, there is not to my knowledge any single defining physical characteristic which the transition to the name 'impeller' implies, which woud not be effectively equivalent to specialised propellers for decelerating ducts.  In systems tending to be described as pumpjets or waterjets hubs tend to be larger, blade-numbers higher, twist, skew and rake lower.  No definitive transition point seems to be accepted as having great significance.  Most commonly literature will refer to a 'rotor', a more general term which could encompass both propellers and impellers.

@@

\subsection{Pump Types} \label{pumptypes}
There are broadly three categories of pump types for water which are used for propulsion.  The first is called an axial-flow. In such a pump, the water has work done on in by blades which push the flow in the same direction to the axis of the blades rotation.  This is what a decelerating ducted propeller comprises. The blades of the rotor push the water from the front, to the back, in the same direction as the craft moves, by blades that turn about an axis that is also oriented in that direction.  These types of pumps are generally best for handling very large flows of water, with relatively less head (energy added) per amount of water flowing through, and are consequently generally desired for propulsion systems, where moving water at high velocity, rather than elevating it to high pressure, is the end objective of the system.

The contrasting system is that of a radial or centrifugal pump.  In one of these systems, work is done on the water by spinning outwards, from close to the axis of rotation to the extremity. In such a case, the axis of the rotor is actually perpendicular to the direction of flow of the water exiting the pump.  The pumps are generally favoured for pump systems which need to add quite a large amount of head, mostly in the form of pressure, to a given amount of flow.  Consequently they are desirable for fire pumps, and or other systems which have to raise water to a high pressure, including lifting water up to great heights. However, centrifugal pumps were originally the type of pump used in the first waterjets created by Hamilton \parencite{hamilton1997}, and are still used in some waterjet applications today, for example in ouboard waterjets \parencite{yamaha2018}.

The third major class of pumps are called mixed-flow pumps, because these types of pumps involve the water having work done on it whilst flowing in both the axial and radial directions.  In overall appearance, they tend to reasonably closely resemble that of an axial flow pump, since the axis of the rotor is parallel to both the direction of water intake and outlet.  However, the water is also accelerated in an outwards direction by the impeller, by means of the hub expanding (and the outer shroud also expanding) in that section where the rotor does work on the water flow. The outwards flow is subsequently deflected back towards the axis by the shape of the shroud contracting towards the nozzle, and the hub tapers to a point in this section also.  The main difference between these can be seen in Figure \ref{fig:AxialMixedHub.png}.

\begin{subfigure}{0.5\textwidth}
\includegraphics[width=0.9\linewidth]{AxialMixed.png}
\caption{}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=0.9\linewidth]{MixedHub.png}
\caption{}
\end{subfigure}

\caption{Mixed flow pumps push the water partly outwards as the water flows through the impeller blades, in a part-way step towards a fully centrifugal or radial flow, as shown in (a) from \cite{carlton2007}. Different hub shapes can induce different levels of radial flow or axial flow in the pump, as seen in (b) from \cite{hamilton1997}}
\label{fig:AxialMixedHub.png}
\end{figure}

The physics of of the fluid flows through the impeller of a mixed-flow pump are quite different to that of a pure-axial pump (the flows must be modeled in three rather than two dimensions), and blade design necessarily changes somewhat to accomodate this type of change.  However, the core limitations and physics surrounding duct design before and after the impeller remain in place for mixed-flow pumpjets, with the primary additional consideration being that the pump section of the duct will tend to need to be somewhat more bulbous in shape to accommodate the larger hub.  As such, differences in the pump efficiency (how well the impellers add energy to the flow) at any particular flow rate might not represent a large impact on the overall efficiency of the propulsor, as the losses due intake duct and nozzle will comprise separate and independent considerations, quite independent of the efficiency of the pump design.

Mixed flow pumps also can experience flow separation or stall in just the same way that this occurs in pure axial-flow pumps \ref{li2013}, and the similar design considerations (such as optimal flow rate) impact the extent to which this occurs. Similarly, the design considerations for reduction of cavitation in mixed-flow pumps tend to be similar, including increasing blade area, and similar efficiency trade-offs are also generally found \parencite{hamilton1997}

Mixed flow pumps can have a variety of different degrees of radial flow over the impeller, and a variety of different designs are found in the pumps in waterjets which vary from pure axial to mixed-flow designs. This is because the optimal efficiency of mixed-flow pumps tend to lie between the optimal efficiencies of axial and centrifugal pumps, across an intermediate range of flow requirements.  Many waterjet manufacturers offer a combination of both pure axial and mixed-flow designs for various applications, and often considerations other than efficiency, including machinery size and shape, can be even more significant than pure efficiency considerations.  Detailed discussions of the differences between pure axial-flow and mixed-flow pumps can be found in \cite{wislicenus1986}.

@@

\subsection{Recent Commentary}
There has been already some attention given to the probable efficiencies of pump jets at low speed. Recently Andrew Davies published a piece in the The Strategist (ASPI) arguing that the pump jet may actually save fuel at transit speeds (around 8kt) and overall improve the endurance of the submarine, despite likely having lower efficiency at low speeds such as patrol speed \parencite{davies2017}.  This argument relies on pumpjet efficiency still overtaking that of propellers at what is considered a relatively low speed for many applications.  As Andrew acknowledges, further quantification of the likely efficiency is required, as the particular chart which he cites is not quantified.  This section aims to identify as far as is reasonably possible from public literature how the efficiency of a likely pumpjet for a submarine might perform in quantitative terms, relative to a propeller.

@@

\begin{figure}[h]
\captionsetup{width=.8\textwidth}
\includegraphics[width=.8\textwidth]{JetHighFlow.png}
\caption{At high flow rates the water has sufficient velocity (indicated by arrow length) to overcome the adverse pressure gradients created by the duct shape, shown by the change from green to red. After the impeller has added further pressure to the water, the water accelerates out of the contracting nozzle through the favourable pressure gradient (red to green) to produce a jet, reaching equal pressure to outside at the exit}
\label{fig:JetHighFlow.png}
\end{figure}

@@

\begin{figure}[h]
\captionsetup{width=.8\textwidth}
\includegraphics[width=.8\textwidth]{JetLowFlow.png}
\caption{At low flow rates the water has insufficient velocity to overcome the adverse pressure gradients created by the duct shape, and flow separation begins to occur on the duct intake, and between the blades beginning near the roots (stall).  Flow through the jet would be unsteady, with excessive energy expended creating vortices and turbulence that results in little thrust.}
\label{fig:JetLowFlow.png}
\end{figure}

@@

Given the fundamental nature of the flow-rate and intake area to the design of pumpjets, and the great lenght of time for which they have been understood and discussed as a crucial design input for a waterjet, it is altogether implausible that some large increase in intake area can be achieved without incurring necessary and unavoidable tradeoffs.
